Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-01-22 paper/current/main.tex
edit the Method section

%
% CONDITIONAL PROBABILITY EQ
%
\begin{equation}
	p(j \given i) = \frac{\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i}, \label{eq:cond_prob_w2v}
\end{equation}
where $\bm{v}$ and $\bm{u}$ are the ``in-vector" and ``out-vector", respectively,  $Z_i=\sum_{x \in \mathcal{A}} \exp(\bm{u}_{x} \cdot \bm{v}_{i})$ is a normalization constant, and $\mathcal{A}$ is the set of all locations.
We follow the standard practice and only use the in-vector, $v$,  which is known to be superior to the out-vector in link prediction benchmarks~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2017gendered}.

Fitting the Eq.~\ref{eq:cond_prob_w2v} are computationally expensive as its denominator involves a sum over all possible entities in data.
Commonly, the negative sampling is introduced as a way to avoid computing the denominator.
The negative sampling is an optimization method widely used for training the machine learning model, which operates as follows.
For each center--context word pair $(i,j)$, one generates $k$ noise center--context word pairs, denoted by $(i, \ell)$, where context word $\ell$ for the noise pair is sampled from a noise distribution, denoted by $p_{0}(\ell)$.
Given a center-context word pair, $(i,j)$, as an input, one predicts whether it comes from the original data or noise distribution based on a similarity between two words, \eg dot similarity in case of the \textit{word2vec}.

Due to the negative sampling, the correct expression optimized with the negative sampling is given by
\begin{align}
	P\left(j \given i \right):= \frac{P(j)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{align}
where $Z_i=\sum_{j' \in \mathcal{A}} P(j') \exp(\bm{u}_{j'} \cdot \bm{v}_{i})$.

@@

\subsection*{Skip-gram and \textit{word2vec}}
\label{sec:sg_and_word2vec}

The \textit{word2vec} takes a form
\begin{align}
	\frac{ f(\bm{u}_{j'}, \bm{v}_i) }{ \sum_{j' \in {\cal A}} f(\bm{u}_{j'}, \bm{v}_i)},
\end{align}
where $f$ is a nonnegative function.
Fitting the probability model is computationally expensive as its denominator involves a sum over all possible entities ${\cal A}$ in data.
To circumvent this problem, the negative sampling solves a binary classification task,
in which one is given an instance $z$ and asked whether $z$ is sampled from an actual data $x$ or a noise data $\tilde x$,
\begin{align}
	\label{eq:sigmoid}
	\frac{1}{1 + \exp\left( -\log f(z) \right)}.
\end{align}
Because this sigmoid function does not involve the normalization constant, the model fitting can be done with less computational burden. Besides, this has a side effect: the negative sampling has a bias and hence does not correctly train the model \cite{Dyer2014}.
The bias can be corrected using the noise contrastive estimation (NCE) \cite{Chia2010}, which solves the
same classification task but with a bias-corrected sigmoid function, \ie
\begin{align}
	\label{eq:sigmoid2}
	\frac{1}{1 + \exp\left( -\log f(z) + \log p_0(z) \right)}.
\end{align}
Note that Eq.~\eqref{eq:sigmoid2} has an additional term $\log p_0(z)$, which is the log-probability of $z$ for a noise distribution.

If the bias is not corrected, what model does the negative sampling actually fit? We rewrite the sigmoid function (Eq.~\eqref{eq:sigmoid}) in form of the bias-corrected sigmoid function, \ie
\begin{align}
	\label{eq:sigmoid3}
	\frac{1}{1 + \exp\left[ - \log \tilde f(z) + \log p_0(z)  \right]},
\end{align}
where we have defined
\begin{align}
	\label{eq:unnormalized}
	\tilde f(z) = f(z) p_0(z).
\end{align}
This sigmoid function correctly fits a probability model:
\begin{align}
	f'(x) / \sum_{x' \in {\cal X}} f'(x),
\end{align}
or, by substituting Eq.~\eqref{eq:unnormalized},
\begin{align}
	p_0(x)f(x) / \sum_{x' \in {\cal X}} p_0(x')f(x').
\end{align}

For the \textit{word2vec}, the correct expression optimized with the negative sampling is given by
\begin{align}
	P\left(j \given i \right):= \frac{p_0(j)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z'_i}.
\end{align}
where $Z'_i=\sum_{x \in \mathcal{A}} p_0(x) \exp(\bm{u}_{x} \cdot \bm{v}_{i})$.

>>>>

%
% CONDITIONAL PROBABILITY EQ
%
\begin{equation}
	p(j \given i) = \frac{\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i}, \label{eq:cond_prob_w2v}
\end{equation}
where $\bm{v}$ and $\bm{u}$ are the ``in-vector" and ``out-vector", respectively,  $Z_i=\sum_{j' \in \mathcal{A}} \exp(\bm{u}_{j'} \cdot \bm{v}_{i})$ is a normalization constant, and $\mathcal{A}$ is the set of all locations.
We follow the standard practice and only use the in-vector, $\bm{v}$,  which is known to be superior to the out-vector in link prediction benchmarks~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2017gendered}.

@@

\subsection*{An implicit bias in the negative sampling}
\label{sec:sg_and_word2vec}

The negative sampling is a biased estimator~\autocite{Chia2010,Dyer2014} that optimizes a model different from the {\it word2vec}.
What is the correct model for the word2vec with the negative sampling?

The negative sampling trains the word2vec by a binary classification task as follows.
One samples a center-context word pair $(i,j)$ from the given data as a positive pair, and $k$ negative pairs $(i,\ell)$.
The negative context $\ell$ is sampled from noise distribution $p_0(\ell)=P^\gamma(\ell)$, where $P(j)$ is the
fraction of $j$ in the given data, and $\gamma>0$ is a parameter.
Then, given a center-context word pair $(i,j)$, one predicts whether it is a positive or negative sample using sigmoid function:
\begin{align}
	\label{eq:sigmoid}
	\frac{1}{1 + \exp\left( - \bm{u}_j \cdot \bm{v}_i \right)}.
\end{align}

The negative sampling does not guaranteed that, even if one increases the number of training samples and iterations, the embedding is not guaranteed to converge to the optimial one~\autocite{Chia2010,Dyer2014}.
This inconsistency problem is addressed by the noise contrastive estimation (NCE) \autocite{Chia2010}, which fits a probability model in form:
\begin{align}
	\label{eq:nce}
	f(x) / \sum_{x'} f(x'),
\end{align}
where $f(x)$ is the unnormalized probability for an instance $x$.
As is the case for the negative sampling, the NCE fits the model based on a binary classification task, with positive and negative samples randomly drawn from the given data and a noise distribution $p_0$, respectively, using a bias-corrected sigmoid function, \ie
\begin{align}
	\label{eq:sigmoid2}
	\frac{1}{1 + \exp\left( -f(x) + \log p_0(x) \right)}.
\end{align}
In case of the word2vec, the instance is the dot similarity $x=\bm{u}_i \cdot \bm{v}_j$,  $f(x)=\exp(x)$, and $p_0(j)=P^\gamma(j)$.

Rewriting the sigmoid function for the negative sampling (Eq.~\eqref{eq:sigmoid}) in form of the bias-corrected sigmoid function yields
\begin{align}
	\label{eq:sigmoid3}
	\frac{1}{1 + \exp\left[ - \tilde f(x) + \log p_0(x)  \right]},
\end{align}
where we have defined
\begin{align}
	\label{eq:unnormalized}
	\tilde f(x) = f(x) p_0(x).
\end{align}
This bias-corrected sigmoid function fits a probability model:
\begin{align}
	f'(x) / \sum_{x'} f'(x),
\end{align}
or, by substituting $x=\bm{u}_i \cdot \bm{v}_j$, $f(x)=\exp(x)$, and $p_0(j)=P^\gamma (j)$,
\begin{align}
	P\left(j \given i \right):= \frac{P^\gamma (j)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i}.
\end{align}
where $Z_i=\sum_{j' \in \mathcal{A}} P^\gamma (j') \exp(\bm{u}_{j'} \cdot \bm{v}_{i})$.
